---
title: "Analysis 1"
author: "James Ashmore"
date: "2019-10-23"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Set knitr chunk options:

```{r knitr}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/01-analysis.Rmd/",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

## Packages

Load required analysis packages:

```{r}
pacman::p_load(batchelor, gam, here, readr, scater, scran, slingshot, RColorBrewer, pheatmap)
```

## Atlas

Load the MouseGastrulationData package:

```{r}
pacman::p_load(MouseGastrulationData)
```

Modify colour palette to include unassigned and unknown cells:

```{r}
EmbryoCelltypeColours <- c(EmbryoCelltypeColours, unassigned = "#999999", unknown = "#999999")
```

Select samples from embryonic stage E8.25:

```{r}
sel <- subset(AtlasSampleMetadata, stage == "E8.25")
```

Obtain processed counts from the dataset:

```{r}
ref <- EmbryoAtlasData(type = "processed", samples = sel$sample)
```

Select cell types from the dataset:

```{r}
sel <- c("Mixed mesoderm", "Cardiomyocytes", "Allantois", "ExE mesoderm", "Mesenchyme", "Haematoendothelial progenitors", "Endothelium")
```

Filter processed counts for select cell types:

```{r}
ref <- ref[, ref$celltype %in% sel]
```

Remove cells with no defined cell type:

```{r}
nan <- is.na(ref$celltype)
ref <- ref[, !nan]
```

Compute log-transformed normalized expression values:

```{r}
ref <- logNormCounts(ref)
```

Change dimension names to those required by the scater package:

```{r}
reducedDimNames(ref) <- c("PCA", "UMAP")
```

Plot UMAP showing all the selected cells: 

```{r}
plt <- plotUMAP(ref, colour_by = "celltype")

ggplot(plt$data, aes(X, Y)) +
  geom_point(aes(colour = colour_by)) + 
  scale_colour_manual(name = "Cell type", values = EmbryoCelltypeColours) +
  theme_void() +
  theme(aspect.ratio = 1)
```

## Allantois

Load the Allantois dataset:

```{r}
sce <- read_rds(here("data/annotation.Rds"))
```

Select wild-type endothelial cells:

```{r}
sce <- sce[, sce$genotype == "Het" & sce$cellassign_celltype == "Endothelial"]
```

## Neighbors

Clear spike-in information across batches:

```{r}
ref <- clearSpikes(ref)

sce <- clearSpikes(sce)
```

Filter batches to contain same genes:

```{r}
rownames(ref) <- rowData(ref)$ENSEMBL

rownames(sce) <- rowData(sce)$gene_id

ids <- intersect(rownames(ref), rownames(sce))

ref <- ref[ids, ]

sce <- sce[ids, ]
```

Select genes with positive average biological components:

```{r}
mod <- model.matrix(~ pool + sample, data = colData(ref))

dec <- combineVar(modelGeneVar(ref, design = mod), modelGeneVar(sce, block = sce$phase))

hvg <- rownames(dec)[dec$bio > 0]
```

Correct for batch effects using a fast version of the MNN method:

```{r}
mnn <- fastMNN(ref, sce, pc.input = TRUE, subset.row = hvg)

colnames(mnn) <- NULL

reducedDimNames(mnn) <- "MNN"
```

Annotate merged dataset with batch and celltype information:

```{r}
mnn$batch <- ifelse(mnn$batch == 1, "Marioni", "Wilson")

mnn$celltype <- c(ref$celltype, sce$cellassign_celltype)
```

Identify kNN from reference cells:

```{r}
sbj <- reducedDim(mnn, "MNN")[mnn$batch == "Marioni", ]

qry <- reducedDim(mnn, "MNN")[mnn$batch == "Wilson", ]

num <- round(sqrt(ncol(ref)))

knn <- BiocNeighbors::queryKNN(sbj, qry, k = num)
```

Annotate cell type using kNN cells:

```{r}
nan <- rep(NA, ncol(sce))

mnn$neighbor <- c(ref$celltype, nan)

lab <- apply(knn$index, 1, function(x) names(which.max(table(ref$celltype[x]))))

mnn$neighbor[mnn$batch == "Wilson"] <- lab
```

Perform UMAP on the PCA data from the combined dataset:

```{r}
set.seed(1701)

mnn <- runUMAP(mnn, n_neighbors = 50, dimred = "MNN")
```

Plot UMAP showing batch annotation:

```{r}
plt <- plotUMAP(mnn, colour_by = "batch")

col <- c("Marioni" = "#d9d9d9", "Wilson" = "#000000")

ggplot(plt$data, aes(X, Y)) +
  geom_point(aes(colour = colour_by)) + 
  scale_colour_manual(name = "Batch", values = col) +
  theme_void() +
  theme(aspect.ratio = 1)
```

Plot UMAP showing celltype annotation:

```{r}
plt <- plotUMAP(mnn)

plt$data$fill_by <- mnn$celltype

plt$data$colour_by <- mnn$batch

col <- c(
  "Allantois" = "#4cb08d",
  "Blood" = "#c04fb4",
  "Cardiac" = "#67b54e",
  "Cardiomyocytes" = "#7b62d0",
  "Endothelial" = "#bcaf46",
  "Endothelium" = "#6479c3",
  "ExE mesoderm" = "#ce5137",
  "Extraembryonic ectoderm" = "#54acd8",
  "Extraembryonic mesoderm" = "#d34379",
  "Gut" = "#667934",
  "Haematoendothelial progenitors" = "#c07ebe",
  "Mesenchyme" = "#c38041",
  "Mesothelium" = "#bd606e"
)

pch <- c("Marioni" = "transparent", "Wilson" = "#000000")

ggplot(plt$data, aes(X, Y)) +
  geom_point(aes(fill = fill_by, colour = colour_by), shape = 21) + 
  scale_fill_manual(name = "Cell type", values = col) + 
  scale_colour_manual(name = "Batch", values = pch, limits = "Wilson") + 
  theme_void() +
  theme(aspect.ratio = 1)
```

Plot UMAP showing neighbour and batch annotation:

```{r}
plt <- plotUMAP(mnn)

plt$data$fill_by <- mnn$neighbor

plt$data$colour_by <- mnn$batch

col <- c(
  "Allantois" = "#4cb08d",
  "Blood" = "#c04fb4",
  "Cardiac" = "#67b54e",
  "Cardiomyocytes" = "#7b62d0",
  "Endothelial" = "#bcaf46",
  "Endothelium" = "#6479c3",
  "ExE mesoderm" = "#ce5137",
  "Extraembryonic ectoderm" = "#54acd8",
  "Extraembryonic mesoderm" = "#d34379",
  "Gut" = "#667934",
  "Haematoendothelial progenitors" = "#c07ebe",
  "Mesenchyme" = "#c38041",
  "Mesothelium" = "#bd606e"
)

pch <- c("Marioni" = "transparent", "Wilson" = "#000000")

ggplot(plt$data, aes(X, Y)) +
  geom_point(aes(fill = fill_by, colour = colour_by), shape = 21) + 
  scale_fill_manual(name = "Neighbor", values = col) + 
  scale_colour_manual(name = "Batch", values = pch, limits = "Wilson") + 
  theme_void() +
  theme(aspect.ratio = 1)
```

Save merged dataset for further exploration:

```{r}
write_rds(mnn, here("data/01-analysis.rds"))
```
