---
title: "Analysis 5"
author: "James Ashmore"
date: "2019-10-23"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Set knitr chunk options:

```{r knitr}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/05-analysis/",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

## Packages

Load required analysis packages:

```{r}
pacman::p_load(batchelor, gam, here, readr, scater, scran, slingshot, RColorBrewer, pheatmap, igraph, EnsDb.Mmusculus.v79)
```

## Import

Load the merged dataset:

```{r}
mnn <- read_rds(here("data/04-analysis.rds"))
```

Annotate gene identifiers:

```{r}
rowData(mnn)$ENSEMBL <- rownames(mnn)

rowData(mnn)$SYMBOL <- mapIds(EnsDb.Mmusculus.v79, keys = rownames(mnn), column = "SYMBOL", keytype = "GENEID")

rownames(mnn) <- uniquifyFeatureNames(rowData(mnn)$ENSEMBL, rowData(mnn)$SYMBOL)
```

Plot UMAP showing neighbour, batch, and genotype annotation:

```{r, fig.width = 10, fig.height = 6}
plt <- plotUMAP(mnn)

plt$data$fill_by <- mnn$neighbor

plt$data$colour_by <- mnn$batch

plt$data$facet_by <- mnn$genotype

col <- c(
  "Allantois" = "#4cb08d",
  "Blood" = "#c04fb4",
  "Cardiac" = "#67b54e",
  "Cardiomyocytes" = "#7b62d0",
  "Endothelial" = "#bcaf46",
  "Endothelium" = "#6479c3",
  "ExE mesoderm" = "#ce5137",
  "Extraembryonic ectoderm" = "#54acd8",
  "Extraembryonic mesoderm" = "#d34379",
  "Gut" = "#667934",
  "Haematoendothelial progenitors" = "#c07ebe",
  "Mesenchyme" = "#c38041",
  "Mesothelium" = "#bd606e"
)

pch <- c("Marioni" = "transparent", "Wilson" = "#000000")

ggplot(plt$data, aes(X, Y)) + 
  geom_point(data = plt$data[, 1:4], mapping = aes(X, Y), colour = "grey") + 
  geom_point(aes(fill = fill_by, colour = colour_by), shape = 21) + 
  facet_wrap(~ facet_by, scales = "fixed") + 
  scale_fill_manual(name = "Neighbor", values = col) + 
  scale_colour_manual(name = "Batch", values = pch, limits = "Wilson") + 
  theme_void() +
  theme(aspect.ratio = 1)
```

## Sub-clustering

```{r}
sclust <- function(obj) {
  
  ref <- mnn[, mnn$batch == "Marioni"]
  
  sce <- mnn[, mnn$batch == "Wilson"]
  
  mod <- model.matrix(~ pool + sample, data = colData(ref))
  
  dec <- combineVar(modelGeneVar(ref, design = mod), modelGeneVar(sce, block = sce$phase))
  
  hvg <- rownames(dec)[dec$bio > 0]
  
  snn <- buildSNNGraph(obj, assay.type = "reconstructed", use.dimred = "MNN", subset.row = hvg)
  
  cls <- igraph::cluster_walktrap(snn)$membership
  
  obj$cluster_walktrap <- factor(cls)
  
  return(obj)
  
}
```

```{r}
tbl <- table(mnn$neighbor)

lvl <- names(tbl)

lvl <- lvl[tbl > 1]
```

```{r}
mnn <- lapply(lvl, function(x) {
  
  obj <- mnn[, mnn$neighbor == x]
  
  obj <- sclust(obj)
  
  obj

})

names(mnn) <- lvl
```

```{r}
sig <- lapply(mnn, function(x) {
  
  m <- findMarkers(x, x$cluster_walktrap, block = x$batch, assay.type = "logcounts")
  
  m <- lapply(m, subset, FDR < 0.1)
  
  m <- lapply(m, head, n = 25)
  
  m <- lapply(m, rownames)
  
  m <- unique(sort(unlist(m)))

})

sig <- Filter(length, sig)
```

### Dimension {.tabset}

Create dimensionality reduction plots coloured by the top 5 marker genes:

```{r, fig.height = 4, fig.width = 6, results = "asis"}
for (n in names(sig)) {
  
  obj <- mnn[[n]]
  
  plt <- plotUMAP(obj, colour_by = "cluster_walktrap", shape_by = "genotype")
  
  cat("#### ", n, "\n")

  print(plt)

  cat("\n\n")

}
```

### Heatmap {.tabset}

Create a heatmap of expression values for the top 5 marker genes:

```{r, fig.height = 8, fig.width = 10, results = "asis"}
for (n in names(sig)) {

  obj <- mnn[[n]]
  
  ids <- sig[[n]]

  cat("#### ", n, "\n")

  plotHeatmap(obj, ids, exprs_values = "reconstructed", center = TRUE, symmetric = TRUE, colour_columns_by = c("batch", "genotype"), order_columns_by = "cluster_walktrap", show_colnames = FALSE, cluster_rows = ifelse(length(ids) >= 2, TRUE, FALSE))

  cat("\n\n")

}
```
