---
title: "Annotation"
author: "James Ashmore"
date: "2019-08-26"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Motivation

In this analysis we are going to annotate previously defined clusters with a cell type label. This process can be done manually using a domain expert's knowledge or automatically based on reference data.

## Setup

Define chunk options:

```{r knitr}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/annotation",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Attach required packages:

```{r pacman}
pacman::p_load(
  cellassign,
  reticulate,
  scater
)
```

Import experiment data:

```{r}
sce <- readr::read_rds(here::here("data/markers.rds"))
```

## Annotation

### Manual {.tabset}

Manually annotate clusters with cell type labels:

```{r, eval = FALSE}
sce$manual_celltype <- "CellType"
```

#### PCA

```{r, eval = FALSE}
plotPCA(sce, colour_by = "manual_celltype")
```

#### t-SNE

```{r, eval = FALSE}
plotTSNE(sce, colour_by = "manual_celltype")
```

#### UMAP

```{r, eval = FALSE}
plotUMAP(sce, colour_by = "manual_celltype")
```

### Automatic {.tabset}

Automatically annotate cells to known types based on the expression patterns of a priori known marker genes:

```{r}
ann <- readr::read_csv(here::here("markers.csv"))

rho <- reshape2::dcast(ann, gene ~ cell, length)

rho <- tibble::column_to_rownames(rho, var = "gene")

rho <- as.matrix(rho)

sel <- intersect(rownames(rho), rownames(sce))

set.seed(1701)

py_set_seed(1701)

fit <- cellassign(sce[sel, ], marker_gene_info = rho[sel, ], s = sizeFactors(sce))

sce$cellassign_celltype <- fit$cell_type
```

#### PCA

```{r}
plotPCA(sce, colour_by = "cellassign_celltype")
```

#### t-SNE

```{r}
plotTSNE(sce, colour_by = "cellassign_celltype")
```

#### UMAP

```{r}
plotUMAP(sce, colour_by = "cellassign_celltype")
```

## Summary

### Output

Write experiment data:

```{r}
readr::write_rds(sce, here::here("data/annotation.Rds"))
```

### Session

Print session information:

```{r session}
devtools::session_info()
```
