---
title: "Clustering"
author: "James Ashmore"
date: "2019-08-14"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Motivation

Clustering is an unsupervised learning procedure that is used in scRNA-seq data analysis to empirically define groups of cells with similar expression profiles. Here, we demonstrate the application of several commonly used clustering methods with our experiment data.

## Setup

Set chunk options:

```{r knitr}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/clustering",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Load required packages:

```{r}
pacman::p_load(
  cluster,
  devtools,
  here,
  igraph,
  pheatmap,
  readr,
  SC3,
  scater,
  scran
)
```

Read experiment data:

```{r}
sce <- read_rds(here("data/dimension.Rds"))
```

## Graph

### Walktrap {.tabset}

Perform clustering using rank-based weights followed by Walktrap clustering:

```{r}
snn <- buildSNNGraph(sce, type = "rank", use.dimred = "PCA")

cls <- cluster_walktrap(snn)$membership

sce$cluster_walktrap <- factor(cls)
```

Calculate the modularity of each cluster:

```{r}
mod <- clusterModularity(snn, sce$cluster_walktrap, as.ratio = TRUE)

mod <- Matrix::forceSymmetric(mod)

pal <- RColorBrewer::brewer.pal(5, "Greens")

col <- colorRampPalette(pal)(10)

brk <- seq(0, max(mod), length.out = 11)

pheatmap(mod, color = col, breaks = brk, cluster_rows = FALSE, cluster_cols = FALSE)
```

Inspect dimensionality reduction plots coloured by explanatory variables:

#### PCA

```{r, fig.height = 8, fig.width = 10}
var <- c("cluster_walktrap", "genotype", "barcodes", "lane", "plate", "phase")

plt <- lapply(var, function(x) plotPCA(sce, colour_by = x))

patchwork::wrap_plots(plt, ncol = 2)
```

#### TSNE

```{r, fig.height = 8, fig.width = 10}
var <- c("cluster_walktrap", "genotype", "barcodes", "lane", "plate", "phase")

plt <- lapply(var, function(x) plotTSNE(sce, colour_by = x))

patchwork::wrap_plots(plt, ncol = 2)
```

#### UMAP

```{r, fig.height = 8, fig.width = 10}
var <- c("cluster_walktrap", "genotype", "barcodes", "lane", "plate", "phase")

plt <- lapply(var, function(x) plotUMAP(sce, colour_by = x))

patchwork::wrap_plots(plt, ncol = 2)
```

### Louvain {.tabset}

Perform clustering using Jaccard-based weights followed by Louvain clustering:

```{r}
snn <- buildSNNGraph(sce, type = "jaccard", use.dimred = "PCA")

cls <- cluster_louvain(snn)$membership

sce$cluster_louvain <- factor(cls)
```

Calculate the modularity of each cluster:

```{r}
mod <- clusterModularity(snn, sce$cluster_louvain, as.ratio = TRUE)

mod <- Matrix::forceSymmetric(mod)

pal <- RColorBrewer::brewer.pal(5, "Greens")

col <- colorRampPalette(pal)(10)

brk <- seq(0, max(mod), length.out = 11)

pheatmap(mod, color = col, breaks = brk, cluster_rows = FALSE, cluster_cols = FALSE)
```

Inspect dimensionality reduction plots coloured by explanatory variables:

#### PCA

```{r, fig.height = 8, fig.width = 10}
var <- c("cluster_louvain", "genotype", "barcodes", "lane", "plate", "phase")

plt <- lapply(var, function(x) plotPCA(sce, colour_by = x))

patchwork::wrap_plots(plt, ncol = 2)
```

#### TSNE

```{r, fig.height = 8, fig.width = 10}
var <- c("cluster_louvain", "genotype", "barcodes", "lane", "plate", "phase")

plt <- lapply(var, function(x) plotTSNE(sce, colour_by = x))

patchwork::wrap_plots(plt, ncol = 2)
```

#### UMAP

```{r, fig.height = 8, fig.width = 10}
var <- c("cluster_louvain", "genotype", "barcodes", "lane", "plate", "phase")

plt <- lapply(var, function(x) plotUMAP(sce, colour_by = x))

patchwork::wrap_plots(plt, ncol = 2)
```

## K-means {.tabset}

Perform k-means clustering on the PCA matrix:

```{r}
set.seed(1701)

dim <- reducedDim(sce, "PCA")

gap <- clusGap(dim, kmeans, K.max = 50)

num <- maxSE(gap$Tab[, "gap"], gap$Tab[, "SE.sim"])

dat <- data.frame(var = seq_len(50), val = gap$Tab[, "gap"])

ggplot(dat, aes(var, val)) +
  geom_point() +
  geom_vline(xintercept = num, colour = "red") +
  labs(x = "Number of clusters", y = "Gap statistic") +
  theme_classic()
```

Select optimal number of clusters:

```{r}
cls <- kmeans(dim, centers = num)

sce$cluster_kmeans <- factor(cls$cluster)
```

Plot the distances between clusters using their centroids:

```{r}
tab <- tabulate(cls$cluster)

dat <- data.frame(wcss = cls$withinss, ncells = tab)

dat$rms <- sqrt(dat$wcss / dat$ncells)

hcl <- hclust(dist(cls$centers), "ward.D2")

plot(hcl)
```

Inspect dimensionality reduction plots coloured by explanatory variables:

### PCA

```{r, fig.height = 8, fig.width = 10}
var <- c("cluster_kmeans", "genotype", "barcodes", "lane", "plate", "phase")

plt <- lapply(var, function(x) plotPCA(sce, colour_by = x))

patchwork::wrap_plots(plt, ncol = 2)
```

### TSNE

```{r, fig.height = 8, fig.width = 10}
var <- c("cluster_kmeans", "genotype", "barcodes", "lane", "plate", "phase")

plt <- lapply(var, function(x) plotTSNE(sce, colour_by = x))

patchwork::wrap_plots(plt, ncol = 2)
```

### UMAP

```{r, fig.height = 8, fig.width = 10}
var <- c("cluster_kmeans", "genotype", "barcodes", "lane", "plate", "phase")

plt <- lapply(var, function(x) plotUMAP(sce, colour_by = x))

patchwork::wrap_plots(plt, ncol = 2)
```

## Hierarchical {.tabset}

Perform hierarchical cluster analysis on the PCA data:

```{r}
dim <- reducedDim(sce, "PCA")

dst <- dist(dim)

cls <- hclust(dst, "ward.D2")

mat <- as.matrix(dst)

cut <- dynamicTreeCut::cutreeDynamic(cls, distM = mat) + 1

sce$cluster_hclust <- factor(cut)
```

Calculate the modularity of each cluster:

```{r}
sil <- silhouette(cut, dist = dst)

plot(sil)
```

Inspect dimensionality reduction plots coloured by clusters:

### PCA

```{r, fig.height = 8, fig.width = 10}
var <- c("cluster_hclust", "genotype", "barcodes", "lane", "plate", "phase")

plt <- lapply(var, function(x) plotPCA(sce, colour_by = x))

patchwork::wrap_plots(plt, ncol = 2)
```

### TSNE

```{r, fig.height = 8, fig.width = 10}
var <- c("cluster_hclust", "genotype", "barcodes", "lane", "plate", "phase")

plt <- lapply(var, function(x) plotTSNE(sce, colour_by = x))

patchwork::wrap_plots(plt, ncol = 2)
```

### UMAP

```{r, fig.height = 8, fig.width = 10}
var <- c("cluster_hclust", "genotype", "barcodes", "lane", "plate", "phase")

plt <- lapply(var, function(x) plotUMAP(sce, colour_by = x))

patchwork::wrap_plots(plt, ncol = 2)
```

## Selection

Choose clustering method for downstream analysis:

```{r}
sce$cluster <- sce$cluster_kmeans
```

## Summary

### Output

Write experiment data:

```{r}
write_rds(sce, here("data/clustering.Rds"))
```

### Session

Print session information:

```{r}
session_info()
```
