---
title: "Doublet detection"
author: "James Ashmore"
date: "2019-10-07"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Motivation

In this analysis we are going to identify and remove doublet libraries so that they do not compromise interpretation of the results.

## Setup

Set chunk options:

```{r knitr}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/doublet",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Load required packages:

```{r}
pacman::p_load(
  devtools,
  here,
  readr,
  scater,
  scran
)
```

Read experiment data:

```{r}
sce <- read_rds(here("data/clustering.Rds"))
```

## By clusters {.tabset}

Identify potential clusters of doublet cells based on intermediate expression profiles:

```{r}
dbl <- doubletCluster(sce, sce$cluster)
```

Inspect output from doublet detection:

```{r}
print(dbl)
```

Find putative marker genes:

```{r}
res <- findMarkers(sce, sce$cluster, block = sce$phase, direction = "up")

sig <- lapply(res, subset, FDR < 0.1)

sig <- Filter(nrow, sig)
```

Plot a heatmap of expression values for cells in the query cluster and the two source clusters:

```{r, fig.height = 8, fig.width = 10, results = "asis"}
for (n in names(sig)) {

  dat <- sig[[n]]
  
  top <- subset(dat, Top <= 10)

  ids <- rownames(top)
  
  id1 <- dbl[rownames(dbl) == n, "source1"]
  
  id2 <- dbl[rownames(dbl) == n, "source2"]
  
  sel <- sce$cluster %in% c(n, id1, id2)

  cat("### Cluster", n, "\n")

  plotHeatmap(sce[, sel], features = ids, x = "cluster", exprs_values = "corrected", colour_columns_by = "cluster", center = TRUE, zlim = c(-1, 1), symmetric = TRUE, cluster_rows = ifelse(length(ids) >= 2, TRUE, FALSE), show_colnames = FALSE)

  cat("\n\n")

}
```

## By simulation {.tabset}

Identify potential doublet cells based on simulations of putative doublet expression profiles:

```{r}
set.seed(1701)

sel <- rowData(sce)$selected

num <- ncol(reducedDim(sce, 'PCA'))

dbl <- doubletCells(sce, subset.row = sel, d = num)

sce$doublet_score <- dbl
```

Plot reduced dimensions coloured by doublet score:

### PCA

```{r}
plotPCA(sce, colour_by = "doublet_score")
```

### TSNE

```{r}
plotTSNE(sce, colour_by = "doublet_score")
```

### UMAP

```{r}
plotUMAP(sce, colour_by = "doublet_score")
```

## Summary

### Output

Write experiment data:

```{r}
write_rds(sce, here("data/doublet.Rds"))
```

### Session

Print session information:

```{r}
session_info()
```
