---
title: "Feature selection"
author: "James Ashmore"
date: "2019-08-14"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

In this document we are going to select features for downstream processing.

## Setup

Set chunk options:

```{r knitr}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/feature",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Load required packages:

```{r pacman}
pacman::p_load(
  eulerr,
  readr,
  pheatmap,
  scales,
  scater,
  scran
)
```

Read object from file:

```{r}
sce <- read_rds(here::here("data/normalization.Rds"))

spk <- grepl("ERCC", rownames(sce))

sce <- splitAltExps(sce, ifelse(spk, "ERCC", "gene"))
```

## Measuring variation

### Variance of expression {.tabset}

Fit a mean-dependent trend to the gene-specific variances:

#### Genes

```{r}
mod <- modelGeneVar(sce)

metadata(sce)$modelGeneVar <- mod

dat <- as.data.frame(mod)

ggplot(dat, aes(mean, total)) +
  geom_point(colour = "grey") +
  geom_line(aes(y = tech)) +
  labs(x = "Mean of log-expression", y = "Variance of log-expression") +
  theme_minimal()
```

#### Spikes

```{r}
mod <- modelGeneVarWithSpikes(sce, "ERCC")

alt <- metadata(mod)

df1 <- as.data.frame(mod)

df2 <- data.frame(mean = alt$mean, total = alt$var)

ggplot(df1, aes(mean, total)) + 
  geom_point(colour = "grey") + 
  geom_line(aes(mean, tech)) + 
  geom_point(data = df2, aes(mean, total), colour = "red") +
  labs(x = "Mean of log-expression", y = "Variance of log-expression") + 
  theme_minimal()
```

#### Poisson

```{r}
mod <- modelGeneVarByPoisson(sce)

dat <- as.data.frame(mod)

ggplot(dat, aes(mean, total)) +
  geom_point(colour = "grey") +
  geom_line(aes(mean, tech)) +
  labs(x = "Mean of log-expression", y = "Variance of log-expression") +
  theme_minimal()
```

### Coefficient of variation {.tabset}

Model the technical coefficient of variation as a function of the mean, and determine the significance of highly variable genes.

#### Genes

```{r}
mod <- modelGeneCV2(sce)

dat <- as.data.frame(mod)

ggplot(dat, aes(mean, total)) +
  geom_point(colour = "grey") +
  geom_line(aes(mean, trend)) +
  scale_x_log10(name = "Mean of expression", labels = comma) +
  scale_y_log10(name = "Coefficient of variation", labels = comma) +
  theme_minimal()
```

#### Spikes

```{r}
mod <- modelGeneCV2WithSpikes(sce, "ERCC")

alt <- metadata(mod)

df1 <- as.data.frame(mod)

df2 <- data.frame(mean = alt$mean, total = alt$cv2)

ggplot(df1, aes(mean, total)) + 
  geom_point(colour = "grey") + 
  geom_point(data = df2, mapping = aes(mean, total), colour = "red") + 
  geom_line(aes(mean, trend)) + 
  scale_x_log10(name = "Mean of expression", labels = comma) + 
  scale_y_log10(name = "Coefficient of variation", labels = comma) + 
  theme_minimal()
```

### Dropout rate

Fit a Michaelis-Menten model to the pattern of dropouts in the data:

```{r}
mat <- normcounts(normalize(sce, return_log = FALSE))

var <- M3Drop::bg__calc_variables(mat)

fit <- M3Drop::bg__fit_MM(var$p, var$s)

dat <- data.frame(mean = var$s, rate = var$p, trend = fit$predictions)

ggplot(dat, aes(mean, rate)) +
  geom_point(colour = "grey") +
  geom_line(aes(mean, trend)) +
  scale_x_log10(labels = comma) +
  labs(x = "Mean of expression", y = "Dropout rate") +
  theme_minimal()
```

## Selecting genes

### Highly variable gene {.tabset}

Model the per-gene variance:

```{r}
mod <- modelGeneVarWithSpikes(sce, "ERCC")

sig <- subset(mod, FDR < 0.01)

sel <- rownames(sig)

metadata(sce)$modelGeneVarWithSpikes <- sel
```

Plot heatmap of gene expression values:

```{r}
plotHeatmap(sce, metadata(sce)$modelGeneVarWithSpikes, center = TRUE, zlim = c(-1, 1), symmetric = TRUE)
```

### Correlated genes

```{r}
mod <- correlatePairs(sce, subset.row = metadata(sce)$modelGeneVarWithSpikes)

sig <- subset(mod, FDR < 0.01)

sel <- union(sig$gene1, sig$gene2)

metadata(sce)$correlatePairs <- sel
```

Plot heatmap of correlated gene expression values:

```{r}
plotHeatmap(sce, metadata(sce)$correlatePairs, center = TRUE, zlim = c(-1, 1), symmetric = TRUE)
```

### High coefficient of variation {.tabset}

```{r}
mod <- modelGeneCV2WithSpikes(sce, "ERCC")

mod$FDR <- p.adjust(mod$p.value, method = "fdr")

sig <- subset(mod, FDR < 0.01)

sel <- rownames(sig)

metadata(sce)$modelGeneCV2WithSpikes <- sel
```

Plot heatmap of genes:

```{r}
plotHeatmap(sce, metadata(sce)$modelGeneCV2WithSpikes, center = TRUE, zlim = c(-1, 1), symmetric = TRUE)
```

### High loading genes {.tabset}

Use principal components analysis to identify high loading genes:

```{r}
mat <- logcounts(sce)

pca <- calculatePCA(t(mat))

var <- attr(pca, "percentVar")

num <- PCAtools::findElbowPoint(var)

len <- seq_len(num)

sum <- rowSums(abs(pca[, len]))

names(sum) <- rownames(sce)

ids <- sort(sum, decreasing = TRUE)

ids <- head(ids, n = 500)

sel <- names(ids)

metadata(sce)$findElbowPoint <- sel
```

Plot heatmap of high loading genes:

```{r}
plotHeatmap(sce, metadata(sce)$findElbowPoint, center = TRUE, zlim = c(-1, 1), symmetric = TRUE)
```

### High dropout genes {.tabset}

Fit Michaelis-Menten curve to model high dropout genes:

```{r}
sce <- normalize(sce, return_log = FALSE)

mat <- normcounts(sce)

fit <- M3Drop::M3DropFeatureSelection(mat, suppress.plot = TRUE)

sel <- fit$Gene

metadata(sce)$M3DropFeatureSelection <- sel
```

Plot heatmap of high dropout genes:

```{r}
plotHeatmap(sce, metadata(sce)$M3DropFeatureSelection, center = TRUE, zlim = c(-1, 1), symmetric = TRUE)
```

## Comparison

### Euler

Compare number of features selected using an Euler diagram:

```{r}
sel <- cbind(
  "All" = rownames(sce) %in% rownames(sce),
  "M3D" = rownames(sce) %in% metadata(sce)$M3DropFeatureSelection,
  "Var" = rownames(sce) %in% metadata(sce)$modelGeneVarWithSpikes,
  "CV2" = rownames(sce) %in% metadata(sce)$modelGeneCV2WithSpikes,
  "PCA" = rownames(sce) %in% metadata(sce)$findElbowPoint,
  "Rho" = rownames(sce) %in% metadata(sce)$correlatePairs
)

fit <- euler(sel)

pal <- RColorBrewer::brewer.pal(n = 5, name = "Pastel1")

lab <- comma(fit$original.values)

plot(fit, fills = pal, quantities = lab)
```

### Jaccard

Calculate Jaccard index:

```{r}
sel <- rbind(
  "M3D" = rownames(sce) %in% metadata(sce)$M3DropFeatureSelection,
  "Var" = rownames(sce) %in% metadata(sce)$modelGeneVarWithSpikes,
  "CV2" = rownames(sce) %in% metadata(sce)$modelGeneCV2WithSpikes,
  "PCA" = rownames(sce) %in% metadata(sce)$findElbowPoint,
  "Rho" = rownames(sce) %in% metadata(sce)$correlatePairs
)

dst <- proxy::dist(sel, method = "Jaccard", diag = TRUE, upper = TRUE)

mat <- as.matrix(dst)

pal <- RColorBrewer::brewer.pal(n = 5, name = "Greens")

col <- colorRampPalette(pal)(10)

brk <- seq(0, 1, length.out = 11)

pheatmap(mat, color = col, breaks = brk, cluster_rows = FALSE, cluster_cols = FALSE)
```

### Dimensionality reduction {.tabset}

```{r}
sel <- c("M3DropFeatureSelection", "modelGeneVarWithSpikes", "modelGeneCV2WithSpikes", "findElbowPoint", "correlatePairs")
```

#### PCA

```{r, fig.height = 11, fig.width = 10}
plt <- lapply(sel, function(x) {
  set.seed(1701)
  row <- metadata(sce)[[x]]
  plotPCA(sce, colour_by = "genotype", rerun = TRUE, run_args = list(subset_row = row)) + ggtitle(x)
})

patchwork::wrap_plots(plt, ncol = 2)
```

#### TSNE

```{r, fig.height = 11, fig.width = 10}
plt <- lapply(sel, function(x) {
  set.seed(1701)
  row <- metadata(sce)[[x]]
  plotTSNE(sce, colour_by = "genotype", rerun = TRUE, run_args = list(subset_row = row)) + ggtitle(x)
})

patchwork::wrap_plots(plt, ncol = 2)
```

#### UMAP

```{r, fig.height = 11, fig.width = 10}
plt <- lapply(sel, function(x) {
  set.seed(1701)
  row <- metadata(sce)[[x]]
  plotUMAP(sce, colour_by = "genotype", rerun = TRUE, run_args = list(subset_row = row)) + ggtitle(x)
})

patchwork::wrap_plots(plt, ncol = 2)
```


## Selection

Select features to use for downstream processing:

```{r}
sel <- rownames(sce) %in% metadata(sce)$modelGeneVarWithSpikes
rowData(sce)$selected <- sel 
```

## Summary

### Output

```{r output}
write_rds(sce, here::here("data/feature.Rds"))
```

### Session

```{r session}
devtools::session_info()
```
