---
title: "Integration"
author: "James Ashmore"
date: "2019-08-23"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Motivation

## Setup

Set chunk options:

```{r knitr}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/integration",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Load required packages:

```{r pacman}
pacman::p_load(
  batchelor,
  devtools,
  here,
  patchwork,
  readr,
  scater,
  scran
)
```

Read experiment data:

```{r readr}
sce <- read_rds(here("data/feature.Rds"))
```

## Diagnosis

### Explanatory variables

Plot explanatory variables ordered by percentage of variance explained:

```{r}
plotExplanatoryVariables(sce, nvars_to_plot = 10)
```

### Dimensionality reduction {.tabset}

Inspect dimensionality reduction plots coloured by explanatory variables:

#### PCA

```{r}
set.seed(1701)

row <- rowData(sce)$selected

sce <- runPCA(sce, subset_row = row)

var <- c("tissue", "genotype", "age", "plate")

plt <- lapply(var, function(x) plotPCA(sce, colour_by = x))

wrap_plots(plt, ncol = 2)
```

#### TSNE

```{r}
set.seed(1701)

row <- rowData(sce)$selected

sce <- runTSNE(sce, subset_row = row)

var <- c("tissue", "genotype", "age", "plate")

plt <- lapply(var, function(x) plotTSNE(sce, colour_by = x))

wrap_plots(plt, ncol = 2)
```

#### UMAP

```{r}
set.seed(1701)

row <- rowData(sce)$selected

sce <- runUMAP(sce, subset_row = row)

var <- c("tissue", "genotype", "age", "plate")

plt <- lapply(var, function(x) plotUMAP(sce, colour_by = x))

wrap_plots(plt, ncol = 2)
```

## Correction

```{r, eval = FALSE}
set.seed(1701)

mnn <- fastMNN(sce, batch = colData(sce)$plate, subset.row = rowData(sce)$selected)

reducedDim(sce, "MNN") <- reducedDim(mnn, "corrected")
```

## Validation

### Dimensionality reduction {.tabset}

Inspect dimensionality reduction plots coloured by explanatory variables:

```{r, eval = FALSE}
row <- rowData(sce)$selected

var <- c("tissue", "genotype", "age", "plate")
```

#### PCA

```{r, eval = FALSE}
plt <- lapply(var, function(x) plotReducedDim(sce, "MNN", colour_by = x))

wrap_plots(plt, ncol = 2)
```

#### TSNE

```{r, eval = FALSE}
set.seed(1701)

sce <- runTSNE(sce, "MNN", subset_row = row)

plt <- lapply(var, function(x) plotTSNE(sce, colour_by = x))

wrap_plots(plt, ncol = 2)
```

#### UMAP

```{r, eval = FALSE}
set.seed(1701)

sce <- runUMAP(sce, "MNN", subset_row = row)

plt <- lapply(var, function(x) plotUMAP(sce, colour_by = x))

wrap_plots(plt, ncol = 2)
```

## Summary

### Output files

Write experiment data:

```{r}
write_rds(sce, here("data/integration.Rds"))
```

### Session information

Print version information:

```{r}
session_info()
```
