---
title: "Integration"
author: "James Ashmore"
date: "2019-08-23"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Motivation

## Setup

Set chunk options:

```{r knitr}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/integration",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Load required packages:

```{r pacman}
pacman::p_load(
  devtools,
  here,
  limma,
  patchwork,
  readr,
  scater,
  scran
)
```

Read experiment data:

```{r readr}
sce <- read_rds(here("data/normalization.Rds"))
```

## Diagnosis

### Explanatory variables

Plot explanatory variables ordered by percentage of variance explained:

```{r}
var <- colData(sce)[, c("genotype", "barcodes", "lane", "plate", "phase")]

plotExplanatoryVariables(sce, nvars_to_plot = 10, variables = var)
```

### Dimensionality reduction {.tabset}

Inspect dimensionality reduction plots coloured by explanatory variables:

```{r}
row <- rowData(sce)$selected

var <- c("genotype", "barcodes", "lane", "plate", "phase")
```

#### PCA

```{r}
set.seed(1701)

reducedDim(sce, "PCA") <- calculatePCA(sce, exprs_values = "logcounts")

plt <- lapply(var, function(x) plotPCA(sce, colour_by = x))

wrap_plots(plt, ncol = 2)
```

#### TSNE

```{r}
set.seed(1701)

reducedDim(sce, "TSNE") <- calculateTSNE(sce, dimred = "PCA")

plt <- lapply(var, function(x) plotTSNE(sce, colour_by = x))

wrap_plots(plt, ncol = 2)
```

#### UMAP

```{r}
set.seed(1701)

reducedDim(sce, "UMAP") <- calculateUMAP(sce, dimred = "PCA")

plt <- lapply(var, function(x) plotUMAP(sce, colour_by = x))

wrap_plots(plt, ncol = 2)
```

## Correction

```{r}
set.seed(1701)

mod <- model.matrix(~ G1 + G2M, sce@colData)

assay(sce, "corrected") <- removeBatchEffect(logcounts(sce), covariates = mod[, -1])
```

## Validation

### Dimensionality reduction {.tabset}

Inspect dimensionality reduction plots coloured by explanatory variables:

```{r}
row <- rowData(sce)$selected

var <- c("genotype", "barcodes", "lane", "plate", "phase")
```

#### PCA

```{r}
set.seed(1701)

reducedDim(sce, "PCA") <- calculatePCA(sce, exprs_values = "corrected")

plt <- lapply(var, function(x) plotPCA(sce, colour_by = x))

wrap_plots(plt, ncol = 2)
```

#### TSNE

```{r}
set.seed(1701)

reducedDim(sce, "TSNE") <- calculateTSNE(sce, dimred = "PCA")

plt <- lapply(var, function(x) plotTSNE(sce, colour_by = x))

wrap_plots(plt, ncol = 2)
```

#### UMAP

```{r}
set.seed(1701)

reducedDim(sce, "UMAP") <- calculateUMAP(sce, dimred = "PCA")

plt <- lapply(var, function(x) plotUMAP(sce, colour_by = x))

wrap_plots(plt, ncol = 2)
```

## Summary

### Output files

Write experiment data:

```{r}
write_rds(sce, here("data/integration.Rds"))
```

### Session information

Print version information:

```{r}
session_info()
```
