---
title: "Normalization"
author: "James Ashmore"
date: "2019-08-14"
output: wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

In this document we are going to compute normalized expression values.

## Setup

Set chunk options:

```{r knitr}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/normalization",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Load required packages:

```{r pacman}
pacman::p_load(
  devtools,
  here,
  knitr,
  patchwork,
  readr,
  reshape2,
  scater,
  scran
)
```

Read experiment data:

```{r}
sce <- read_rds(here("data/quality.Rds"))
```

## Methods

Evaluate different normalization methods using a copy of the experiment data:

```{r}
new <- sce
```

### Raw {.tabset}

Compute expression values from raw counts without any normalization:

```{r raw-normalize}
set.seed(1701)

sizeFactors(new) <- 1

new <- computeSpikeFactors(new, general.use = FALSE)

new <- logNormCounts(new)

new <- runPCA(new)

new <- runTSNE(new)

new <- runUMAP(new)
```

Inspect relative log expression plots coloured by column data:

```{r, fig.height = 5, fig.width = 10}
var <- c("tissue", "genotype")

plt <- lapply(var, function(x) plotRLE(new, colour_by = x))

wrap_plots(plt, ncol = 2)
```

Inspect dimensionality reduction plots coloured by QC metrics:

#### PCA

```{r raw-pca, fig.height = 8, fig.width = 10}
var <- c("tissue", "genotype", "log10_total_counts", "log10_total_features_by_counts", "pct_counts_ERCC", "pct_counts_MT")

plt <- lapply(var, function(x) plotPCA(new, colour_by = x))

wrap_plots(plt, ncol = 2)
```

#### TSNE

```{r raw-tsne, fig.height = 8, fig.width = 10}
var <- c("tissue", "genotype", "log10_total_counts", "log10_total_features_by_counts", "pct_counts_ERCC", "pct_counts_MT")

plt <- lapply(var, function(x) plotTSNE(new, colour_by = x))

wrap_plots(plt, ncol = 2)
```

#### UMAP

```{r raw-umap, fig.height = 8, fig.width = 10}
var <- c("tissue", "genotype", "log10_total_counts", "log10_total_features_by_counts", "pct_counts_ERCC", "pct_counts_MT")

plt <- lapply(var, function(x) plotUMAP(new, colour_by = x))

wrap_plots(plt, ncol = 2)
```


### Library {.tabset}

Compute expression values from raw counts by scaling with library size factors:

```{r library-normalize}
set.seed(1701)

sizeFactors(new) <- librarySizeFactors(new)

new <- computeSpikeFactors(new, general.use = FALSE)

new <- logNormCounts(new)

new <- runPCA(new)

new <- runTSNE(new)

new <- runUMAP(new)
```

Inspect relative log expression plots coloured by column data:

```{r, fig.height = 5, fig.width = 10}
var <- c("tissue", "genotype")

plt <- lapply(var, function(x) plotRLE(new, colour_by = x))

wrap_plots(plt, ncol = 2)
```

Inspect dimensionality reduction plots coloured by QC metrics:

#### PCA

```{r library-pca, fig.height = 8, fig.width = 10}
var <- c("tissue", "genotype", "log10_total_counts", "log10_total_features_by_counts", "pct_counts_ERCC", "pct_counts_MT")

plt <- lapply(var, function(x) plotPCA(new, colour_by = x))

wrap_plots(plt, ncol = 2)
```

#### TSNE

```{r library-tsne, fig.height = 8, fig.width = 10}
var <- c("tissue", "genotype", "log10_total_counts", "log10_total_features_by_counts", "pct_counts_ERCC", "pct_counts_MT")

plt <- lapply(var, function(x) plotTSNE(new, colour_by = x))

wrap_plots(plt, ncol = 2)
```

#### UMAP

```{r library-umap, fig.height = 8, fig.width = 10}
var <- c("tissue", "genotype", "log10_total_counts", "log10_total_features_by_counts", "pct_counts_ERCC", "pct_counts_MT")

plt <- lapply(var, function(x) plotUMAP(new, colour_by = x))

wrap_plots(plt, ncol = 2)
```

### Deconvolution {.tabset}

Compute expression values from raw counts by scaling with "deconvolved" size factors:

```{r deconvolution-normalize}
set.seed(1701)

cls <- quickCluster(new, use.ranks = FALSE)

new <- computeSumFactors(new, cluster = cls, min.mean = 1)

new <- computeSpikeFactors(new, general.use = FALSE)

new <- logNormCounts(new)

new <- runPCA(new)

new <- runTSNE(new)

new <- runUMAP(new)
```

Inspect relative log expression plots coloured by column data:

```{r, fig.height = 5, fig.width = 10}
var <- c("tissue", "genotype")

plt <- lapply(var, function(x) plotRLE(new, colour_by = x))

wrap_plots(plt, ncol = 2)
```

Inspect dimensionality reduction plots coloured by QC metrics:

#### PCA

```{r deconvolution-pca, fig.height = 8, fig.width = 10}
var <- c("tissue", "genotype", "log10_total_counts", "log10_total_features_by_counts", "pct_counts_ERCC", "pct_counts_MT")

plt <- lapply(var, function(x) plotPCA(new, colour_by = x))

wrap_plots(plt, ncol = 2)
```

#### TSNE

```{r deconvolution-tsne, fig.height = 8, fig.width = 10}
var <- c("tissue", "genotype", "log10_total_counts", "log10_total_features_by_counts", "pct_counts_ERCC", "pct_counts_MT")

plt <- lapply(var, function(x) plotTSNE(new, colour_by = x))

wrap_plots(plt, ncol = 2)
```

#### UMAP

```{r deconvolution-umap, fig.height = 8, fig.width = 10}
var <- c("tissue", "genotype", "log10_total_counts", "log10_total_features_by_counts", "pct_counts_ERCC", "pct_counts_MT")

plt <- lapply(var, function(x) plotUMAP(new, colour_by = x))

wrap_plots(plt, ncol = 2)
```

### Spike {.tabset}

Compute expression values from raw counts by scaling with spike-in size factors:

```{r spike-normalize}
set.seed(1701)

new <- computeSpikeFactors(new, general.use = TRUE)

new <- logNormCounts(new)

new <- runPCA(new)

new <- runTSNE(new)

new <- runUMAP(new)
```

Inspect relative log expression plots coloured by column data:

```{r, fig.height = 5, fig.width = 10}
var <- c("tissue", "genotype")

plt <- lapply(var, function(x) plotRLE(new, colour_by = x))

wrap_plots(plt, ncol = 2)
```

Inspect dimensionality reduction plots coloured by QC metrics:

#### PCA

```{r spike-pca, fig.height = 8, fig.width = 10}
var <- c("tissue", "genotype", "log10_total_counts", "log10_total_features_by_counts", "pct_counts_ERCC", "pct_counts_MT")

plt <- lapply(var, function(x) plotPCA(new, colour_by = x))

wrap_plots(plt, ncol = 2)
```

#### TSNE

```{r spike-tsne, fig.height = 8, fig.width = 10}
var <- c("tissue", "genotype", "log10_total_counts", "log10_total_features_by_counts", "pct_counts_ERCC", "pct_counts_MT")

plt <- lapply(var, function(x) plotTSNE(new, colour_by = x))

wrap_plots(plt, ncol = 2)
```

#### UMAP

```{r spike-umap, fig.height = 8, fig.width = 10}
var <- c("tissue", "genotype", "log10_total_counts", "log10_total_features_by_counts", "pct_counts_ERCC", "pct_counts_MT")

plt <- lapply(var, function(x) plotUMAP(new, colour_by = x))

wrap_plots(plt, ncol = 2)
```

### Downsample {.tabset}

Compute expression values from raw counts by downsampling:

```{r downsample-normalize}
set.seed(1701)

lib <- librarySizeFactors(new)

fct <- min(lib) / lib

counts(new) <- DropletUtils::downsampleMatrix(counts(new), prop = fct)

sizeFactors(new) <- librarySizeFactors(new)

new <- logNormCounts(new)

new <- runPCA(new)

new <- runTSNE(new)

new <- runUMAP(new)
```

Inspect relative log expression plots coloured by column data:

```{r, fig.height = 5, fig.width = 10}
var <- c("tissue", "genotype")

plt <- lapply(var, function(x) plotRLE(new, colour_by = x))

wrap_plots(plt, ncol = 2)
```

Inspect dimensionality reduction plots coloured by QC metrics:

#### PCA

```{r downsample-pca, fig.height = 8, fig.width = 10}
var <- c("tissue", "genotype", "log10_total_counts", "log10_total_features_by_counts", "pct_counts_ERCC", "pct_counts_MT")

plt <- lapply(var, function(x) plotPCA(new, colour_by = x))

wrap_plots(plt, ncol = 2)
```

#### TSNE

```{r downsample-tsne, fig.height = 8, fig.width = 10}
var <- c("tissue", "genotype", "log10_total_counts", "log10_total_features_by_counts", "pct_counts_ERCC", "pct_counts_MT")

plt <- lapply(var, function(x) plotTSNE(new, colour_by = x))

wrap_plots(plt, ncol = 2)
```

#### UMAP

```{r downsample-umap, fig.height = 8, fig.width = 10}
var <- c("tissue", "genotype", "log10_total_counts", "log10_total_features_by_counts", "pct_counts_ERCC", "pct_counts_MT")

plt <- lapply(var, function(x) plotUMAP(new, colour_by = x))

wrap_plots(plt, ncol = 2)
```

## Selection

Apply the chosen normalization to the original experiment data:

```{r}
set.seed(1701)

cls <- quickCluster(sce, use.ranks = FALSE)

sce <- computeSumFactors(sce, cluster = cls, min.mean = 1)

sce <- computeSpikeFactors(sce, general.use = FALSE)

sce <- logNormCounts(sce)
```

## Summary

### Output

```{r output}
write_rds(sce, here("data/normalization.Rds"))
```

## Session

Print version information:

```{r session}
session_info()
```
