---
title: "Reference mapping"
author: "James Ashmore"
date: "2019-09-18"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Set chunk options:

```{r knitr}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/reference",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Load required packages:

```{r pacman}
pacman::p_load(batchelor, ggalluvial, here, readr, scater, scran, scmap)
```

Read experiment data:

```{r}
sce <- read_rds(here("data/trajectory.Rds"))
```

Select samples from genotype Het:

```{r}
tmp <- sce[, sce$genotype == "Het"]
```

## Reference

### Installation

Install the MouseGastrulationData package:

```{r, eval = FALSE}
BiocManager::install("MouseGastrulationData")
```

Load the MouseGastrulationData package:

```{r}
pacman::p_load(MouseGastrulationData)
```

Modify colour palette to include unassigned cells:

```{r}
EmbryoCelltypeColours <- c(EmbryoCelltypeColours, "unassigned" = "#999999", "unknown" = "#999999")
```

### Selection

Select samples from embryonic stage E8.25:

```{r}
sel <- subset(AtlasSampleMetadata, stage == "E8.25")
```

Obtain processed counts from the mouse gastrulation dataset:

```{r}
ref <- EmbryoAtlasData(type = "processed", samples = sel$sample)
```

### Modification

Change dimension names to those required by the scater package:

```{r}
reducedDimNames(ref) <- c("PCA", "UMAP")
```

Compute log-transformed normalized expression values:

```{r}
ref <- logNormCounts(ref)
```

Remove cells with no defined cell type:

```{r}
nan <- is.na(ref$celltype)
ref <- ref[, !nan]
```

### Exploration

Inspect PCA coloured by cell type:

```{r, fig.height = 6, fig.width = 10}
plt <- plotPCA(ref, colour_by = "celltype")

ggplot(plt$data, aes(X, Y, colour = colour_by)) +
  geom_point(size = 1) +
  scale_colour_manual(name = "Cell type", values = EmbryoCelltypeColours) +
  guides(colour = guide_legend(override.aes = list(size = 2))) +
  theme_void()
```

Inspect UMAP coloured by cell type:

```{r, fig.height = 6, fig.width = 10}
plt <- plotUMAP(ref, colour_by = "celltype")

ggplot(plt$data, aes(X, Y, colour = colour_by)) +
  geom_point(size = 1) +
  scale_colour_manual(name = "Cell type", values = EmbryoCelltypeColours) +
  guides(colour = guide_legend(override.aes = list(size = 2))) +
  theme_void()
```

## Projection

### By cluster

Create feature_symbol column required by scmap package:

```{r}
rowData(tmp)$feature_symbol <- uniquifyFeatureNames(rowData(tmp)$gene_id, rowData(tmp)$gene_name)

rownames(tmp) <- rowData(tmp)$feature_symbol

rowData(ref)$feature_symbol <- uniquifyFeatureNames(rowData(ref)$ENSEMBL, rowData(ref)$SYMBOL)

rownames(ref) <- rowData(ref)$feature_symbol
```

Subset both datasets to common universe of genes:

```{r}
ids <- intersect(rownames(tmp), rownames(ref))

tmp <- tmp[ids, ]

ref <- ref[ids, ]
```

Select most informative features from the reference dataset:

```{r}
ref <- selectFeatures(ref, n_features = 2000)
```

Create a cluster index based on celltype:

```{r}
ref <- indexCluster(ref, cluster_col = "celltype")
```

Project our dataset to the reference dataset:

```{r}
fit <- scmapCluster(projection = tmp, index_list = list(reference = metadata(ref)$scmap_cluster_index))

tmp$scmap_cluster_labs <- fit$scmap_cluster_labs[, "reference"]
```

Plot the assignment of projected cell types:

```{r}
dat <- data.frame(cluster = tmp$cluster, celltype = tmp$scmap_cluster_labs)

dat <- plyr::count(dat, vars = c("cluster", "celltype"))

ggplot(dat, aes(y = freq, axis1 = cluster, axis2 = celltype)) +
  geom_alluvium(aes(fill = celltype)) +
  geom_stratum() +
  geom_text(stat = "stratum", label.strata = TRUE) +
  scale_x_discrete(limits = c("Cluster", "Cell type"), expand = c(0.05, 0.05)) +
  scale_fill_manual(name = "Cell type", values = EmbryoCelltypeColours) +
  theme_minimal()
```

Inspect the UMAP coordinates coloured by projected cell type:

```{r}
plt <- plotUMAP(tmp, colour_by = "scmap_cluster_labs")

ggplot(plt$data, aes(X, Y, colour = colour_by)) +
  geom_point(size = 1) +
  scale_colour_manual(name = "Cell type", values = EmbryoCelltypeColours) +
  guides(colour = guide_legend(override.aes = list(size = 2))) +
  theme_void()
```

Save projected cell types to unmodified experiment data:

```{r}
sce$scmapCluster_celltype <- NA

sce$scmapCluster_celltype[sce$genotype == "Het"] <- tmp$scmap_cluster_labs
```

### By cell

Create a cell index:

```{r}
set.seed(1701)

ref <- indexCell(ref)
```

Project our dataset to the reference dataset:

```{r}
num <- round(sqrt(ncol(ref)))

fit <- scmapCell(projection = tmp, index_list = list(reference = metadata(ref)$scmap_cell_index), w = num)

lab <- apply(fit$reference$cells, 2, function(x) names(which.max(table(ref$celltype[x]))))

tmp$scmap_cell_labs <- lab
```

Plot the assignment of projected cell types:

```{r}
dat <- data.frame(cluster = tmp$cluster, celltype = tmp$scmap_cell_labs)

dat <- plyr::count(dat, vars = c("cluster", "celltype"))

ggplot(dat, aes(y = freq, axis1 = cluster, axis2 = celltype)) +
  geom_alluvium(aes(fill = celltype)) +
  geom_stratum() +
  geom_text(stat = "stratum", label.strata = TRUE) +
  scale_x_discrete(limits = c("Cluster", "Cell type"), expand = c(0.05, 0.05)) +
  scale_fill_manual(name = "Cell type", values = EmbryoCelltypeColours) +
  theme_minimal()
```

Inspect the UMAP coordinates coloured by projected cell type:

```{r}
plt <- plotUMAP(tmp, colour_by = "scmap_cell_labs")

ggplot(plt$data, aes(X, Y, colour = colour_by)) +
  geom_point(size = 1) +
  scale_colour_manual(name = "Cell type", values = EmbryoCelltypeColours) +
  guides(colour = guide_legend(override.aes = list(size = 2))) +
  theme_void()
```

Save projected cell types to unmodified experiment data:

```{r}
sce$scmapCell_celltype <- NA

sce$scmapCell_celltype[sce$genotype == "Het"] <- tmp$scmap_cell_labs
```

## Neighbors

Clear spike-in information across batches:

```{r}
ref <- clearSpikes(ref)

tmp <- clearSpikes(tmp)
```

Select genes with positive average biological components:

```{r}
mod <- model.matrix(~ pool + sample, data = colData(ref))

dec <- combineVar(modelGeneVar(ref, design = mod), modelGeneVar(tmp, block = tmp$phase))

hvg <- rownames(dec)[dec$bio > 0]
```

Correct for batch effects using a fast version of the MNN method:

```{r}
mnn <- fastMNN(ref, tmp, pc.input = TRUE, subset.row = hvg)

reducedDimNames(mnn) <- "MNN"

mnn$batch <- ifelse(mnn$batch == 1, "Marioni", "Wilson")
```

Identify kNN from reference cells:

```{r}
sbj <- reducedDim(mnn, "MNN")[mnn$batch == "Marioni", ]

qry <- reducedDim(mnn, "MNN")[mnn$batch == "Wilson", ]

num <- round(sqrt(ncol(ref)))

knn <- BiocNeighbors::queryKNN(sbj, qry, k = num)
```

Annotate cell type using kNN cells:

```{r}
mnn$celltype <- c(ref$celltype, rep(NA, ncol(tmp)))

lab <- apply(knn$index, 1, function(x) names(which.max(table(ref$celltype[x]))))

mnn$celltype[mnn$batch == "Wilson"] <- lab

tmp$mnn_celltype <- lab
```

Plot the assignment of projected cell types:

```{r}
dat <- data.frame(cluster = tmp$cluster, celltype = tmp$mnn_celltype)

dat <- plyr::count(dat, vars = c("cluster", "celltype"))

ggplot(dat, aes(y = freq, axis1 = cluster, axis2 = celltype)) +
  geom_alluvium(aes(fill = celltype)) +
  geom_stratum() +
  geom_text(stat = "stratum", label.strata = TRUE) +
  scale_x_discrete(limits = c("Cluster", "Cell type"), expand = c(0.05, 0.05)) +
  scale_fill_manual(name = "Cell type", values = EmbryoCelltypeColours) +
  theme_minimal()
```

Plot UMAP of experiment data coloured by cell type annotation:

```{r}
plt <- plotUMAP(tmp, colour_by = "mnn_celltype")

ggplot(plt$data, aes(X, Y, colour = colour_by)) +
  geom_point(size = 1) +
  scale_colour_manual(name = "Cell type", values = EmbryoCelltypeColours) +
  guides(colour = guide_legend(override.aes = list(size = 2))) +
  theme_void()
```

Perform UMAP on the PCA data from the combined dataset:

```{r}
set.seed(1701)

mnn <- runUMAP(mnn, dimred = "MNN")
```

Plot UMAP of combined dataset:

```{r, fig.height = 6, fig.width = 10}
plt <- plotUMAP(mnn, colour_by = "celltype", shape_by = "batch")

ggplot(subset(plt$data, shape_by == "Marioni"), aes(X, Y, colour = colour_by)) +
  geom_point(alpha = 0.5, size = 0.5) +
  scale_colour_manual(values = EmbryoCelltypeColours) +
  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  geom_point(data = subset(plt$data, shape_by == "Wilson"), aes(X, Y), colour = "black", shape = 4) +
  theme_void()
```

Save projected celltype to unmodified experiment data:

```{r}
sce$fastMNN_celltype <- NA

sce$fastMNN_celltype[sce$genotype == "Het"] <- tmp$mnn_celltype
```

## Summary

### Output

Write experiment data:

```{r}
write_rds(sce, here("data/reference.Rds"))
```

### Session

Print session information:

```{r}
devtools::session_info()
```

