---
title: "Reference mapping"
author: "James Ashmore"
date: "2019-09-18"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Set chunk options:

```{r knitr}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/reference",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Load required packages:

```{r pacman}
pacman::p_load(batchelor, ggalluvial, here, pheatmap, readr, scater, scran, scmap, viridis)
```

Read experiment data:

```{r}
sce <- read_rds(here("data/trajectory.Rds"))
```

Select samples from genotype Het:

```{r}
sce <- sce[, sce$genotype == "Het"]
```

Import colour palette for cell types:

```{r}
pal <- read_csv(here("palette.csv"))

col <- pal$colour

names(col) <- pal$celltype
```

## Reference

### Installation

Install the MouseGastrulationData package:

```{r, eval = FALSE}
BiocManager::install("MouseGastrulationData")
```

Load the MouseGastrulationData package:

```{r}
pacman::p_load(MouseGastrulationData)
```

### Selection

Select samples from embryonic stage E8.25:

```{r}
sel <- subset(AtlasSampleMetadata, stage == "E8.25")
```

Obtain processed counts from the mouse gastrulation dataset:

```{r}
ref <- EmbryoAtlasData(type = "processed", samples = sel$sample)
```

### Modification

Change dimension names to those required by the scater package:

```{r}
reducedDimNames(ref) <- c("PCA", "UMAP")
```

Compute log-transformed normalized expression values:

```{r}
ref <- logNormCounts(ref)
```

Remove cells with no defined cell type:

```{r}
nan <- is.na(ref$celltype)
ref <- ref[, !nan]
```

### Exploration

Inspect PCA coloured by cell type:

```{r, fig.height = 6, fig.width = 10}
plt <- plotPCA(ref, colour_by = "celltype")

ggplot(plt$data, aes(X, Y, colour = colour_by)) +
  geom_point(size = 1) +
  scale_colour_manual(name = "Cell type", values = col) +
  guides(colour = guide_legend(override.aes = list(size = 2))) +
  theme_void()
```

Inspect UMAP coloured by cell type:

```{r, fig.height = 6, fig.width = 10}
plt <- plotUMAP(ref, colour_by = "celltype")

ggplot(plt$data, aes(X, Y, colour = colour_by)) +
  geom_point(size = 1) +
  scale_colour_manual(name = "Cell type", values = col) +
  guides(colour = guide_legend(override.aes = list(size = 2))) +
  theme_void()
```

## Projection

### By cluster

Create feature_symbol column required by scmap package:

```{r}
rowData(sce)$feature_symbol <- uniquifyFeatureNames(rowData(sce)$gene_id, rowData(sce)$gene_name)

rownames(sce) <- rowData(sce)$feature_symbol

rowData(ref)$feature_symbol <- uniquifyFeatureNames(rowData(ref)$ENSEMBL, rowData(ref)$SYMBOL)

rownames(ref) <- rowData(ref)$feature_symbol
```

Subset both datasets to common universe of genes:

```{r}
ids <- intersect(rownames(sce), rownames(ref))

sce <- sce[ids, ]

ref <- ref[ids, ]
```

Select most informative features from the reference dataset:

```{r}
ref <- selectFeatures(ref, n_features = 2000)
```

Create a cluster index based on celltype:

```{r}
ref <- indexCluster(ref, cluster_col = "celltype")
```

Inspect the cell type index:

```{r}
idx <- metadata(ref)$scmap_cluster_index

mat <- as.matrix(idx)

brk <- seq(0, max(mat), length.out = 101)

pheatmap(mat, color = viridis(100), breaks = brk, scale = "row")
```

Project our dataset to the reference dataset:

```{r}
fit <- scmapCluster(projection = sce, index_list = list(reference = metadata(ref)$scmap_cluster_index))

sce$scmap_cluster_labs <- fit$scmap_cluster_labs[, "reference"]
```

Plot the assignment of projected cell types:

```{r}
dat <- data.frame(cluster = sce$cluster, celltype = sce$scmap_cluster_labs)

dat <- plyr::count(dat, vars = c("cluster", "celltype"))

ggplot(dat, aes(y = freq, axis1 = cluster, axis2 = celltype)) +
  geom_alluvium(aes(fill = celltype)) +
  geom_stratum() +
  geom_text(stat = "stratum", label.strata = TRUE) +
  scale_x_discrete(limits = c("Cluster", "Cell type"), expand = c(0.05, 0.05)) +
  scale_fill_manual(name = "Cell type", values = col) +
  theme_minimal()
```

Inspect the UMAP coordinates coloured by projected cell type:

```{r}
plt <- plotUMAP(sce, colour_by = "scmap_cluster_labs")

ggplot(plt$data, aes(X, Y, colour = colour_by)) +
  geom_point(size = 1) +
  scale_colour_manual(name = "Cell type", values = col) +
  guides(colour = guide_legend(override.aes = list(size = 2))) +
  theme_void()
```

### By cell

Create a cell index:

```{r}
set.seed(1701)

ref <- indexCell(ref)
```

Project our dataset to the reference dataset:

```{r}
num <- round(sqrt(length(unique(ref$celltype))))

fit <- scmapCell(projection = sce, index_list = list(reference = metadata(ref)$scmap_cell_index), w = num)

lab <- apply(fit$reference$cells, 2, function(x) names(which.max(table(ref$celltype[x]))))

sce$scmap_cell_labs <- lab
```

Plot the assignment of projected cell types:

```{r}
dat <- data.frame(cluster = sce$cluster, celltype = sce$scmap_cell_labs)

dat <- plyr::count(dat, vars = c("cluster", "celltype"))

ggplot(dat, aes(y = freq, axis1 = cluster, axis2 = celltype)) +
  geom_alluvium(aes(fill = celltype)) +
  geom_stratum() +
  geom_text(stat = "stratum", label.strata = TRUE) +
  scale_x_discrete(limits = c("Cluster", "Cell type"), expand = c(0.05, 0.05)) +
  scale_fill_manual(name = "Cell type", values = col) +
  theme_minimal()
```

Inspect the UMAP coordinates coloured by projected cell type:

```{r}
plt <- plotUMAP(sce, colour_by = "scmap_cell_labs")

ggplot(plt$data, aes(X, Y, colour = colour_by)) +
  geom_point(size = 1) +
  scale_colour_manual(name = "Cell type", values = col) +
  guides(colour = guide_legend(override.aes = list(size = 2))) +
  theme_void()
```

## Neighbors

Clear spike-in information across batches:

```{r}
ref <- clearSpikes(ref)

sce <- clearSpikes(sce)
```

Select genes with positive average biological components:

```{r}
mod <- model.matrix(~ pool + sample, data = colData(ref))

dec <- combineVar(modelGeneVar(ref, design = mod), modelGeneVar(sce))

hvg <- rownames(dec)[dec$bio > 0]
```

Correct for batch effects using a fast version of the MNN method:

```{r}
mnn <- fastMNN(ref, sce, pc.input = TRUE, subset.row = hvg)

reducedDimNames(mnn) <- "MNN"
```

Add metadata to the combined dataset:

```{r}
mnn$celltype <- c(ref$celltype, rep("unknown", ncol(sce)))

mnn$pool <- c(ref$pool, rep("unknown", ncol(sce)))

mnn$sample <- c(ref$sample, rep("unknown", ncol(sce)))
```

Annotate experiment cells using kNN coordinates from reference cells:

```{r}
sbj <- reducedDim(mnn, "MNN")[mnn$batch == 1, ]

qry <- reducedDim(mnn, "MNN")[mnn$batch == 2, ]

num <- round(sqrt(length(unique(ref$celltype))))

knn <- BiocNeighbors::queryKNN(sbj, qry, k = num)

lab <- apply(knn$index, 1, function(x) names(which.max(table(ref$celltype[x]))))

mnn$celltype[mnn$batch == 2] <- lab

sce$mnn_cell_labs <- lab
```

Plot the assignment of projected cell types:

```{r}
dat <- data.frame(cluster = sce$cluster, celltype = sce$mnn_cell_labs)

dat <- plyr::count(dat, vars = c("cluster", "celltype"))

ggplot(dat, aes(y = freq, axis1 = cluster, axis2 = celltype)) +
  geom_alluvium(aes(fill = celltype)) +
  geom_stratum() +
  geom_text(stat = "stratum", label.strata = TRUE) +
  scale_x_discrete(limits = c("Cluster", "Cell type"), expand = c(0.05, 0.05)) +
  scale_fill_manual(name = "Cell type", values = col) +
  theme_minimal()
```

Plot UMAP of experiment data coloured by cell type annotation:

```{r}
plt <- plotUMAP(sce, colour_by = "mnn_cell_labs")

ggplot(plt$data, aes(X, Y, colour = colour_by)) +
  geom_point(size = 1) +
  scale_colour_manual(name = "Cell type", values = col) +
  guides(colour = guide_legend(override.aes = list(size = 2))) +
  theme_void()
```

Perform UMAP on the PCA data from the combined dataset:

```{r}
set.seed(1701)

mnn <- runUMAP(mnn, dimred = "MNN")
```

Plot UMAP of combined dataset:

```{r, fig.height = 6, fig.width = 10}
plt <- plotUMAP(mnn, colour_by = "celltype", shape_by = "batch")

ggplot(subset(plt$data, shape_by == 1), aes(X, Y, colour = colour_by)) +
  geom_point(alpha = 0.5, size = 0.5) +
  scale_colour_manual(values = col) +
  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  geom_point(data = subset(plt$data, shape_by == 2), aes(X, Y), colour = "black", shape = 4) +
  theme_void()
```

## Summary

### Output

Write combined dataset:

```{r}
write_rds(mnn, here("data/reference.Rds"))
```

### Session

Print session information:

```{r}
devtools::session_info()
```

