---
title: "Trajectory"
author: "James Ashmore"
date: "2019-08-23"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Motivation

In this analysis we are going to infer trajectories between the clusters or cell types defined previously.

## Setup

Set chunk options:

```{r knitr}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/trajectory",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Load required packages:

```{r pacman}
pacman::p_load(
  SingleCellExperiment,
  slingshot
)
```

Read experiment data:

```{r}
sce <- readr::read_rds(here::here("data/comparison.Rds"))
```

## Slingshot

### Clusters

Infer trajectory between all clusters using UMAP dimensions:

```{r}
sce <- slingshot(sce, clusterLabels = "cluster", reducedDim = "UMAP")
```

Extract slingshot results from experiment data:

```{r}
res <- SlingshotDataSet(sce)
```

Inspect UMAP plot coloured by clustering:

```{r}
pal <- RColorBrewer::brewer.pal(9, name = "Set1")

len <- length(unique(sce$cluster))

pal <- pal[1:len]

names(pal) <- unique(sce$cluster)

col <- pal[match(sce$cluster, names(pal))]

plot(res@reducedDim, col = col, pch = 16, asp = 1, xlab = "Dimension 1", ylab = "Dimension 2")

lines(res, type = "lineages")

legend("topright", legend = unique(sce$cluster), col = pal, pch = 16)
```

Inspect UMAP plot coloured by genotype:

```{r}
pal <- RColorBrewer::brewer.pal(9, name = "Set1")

len <- length(unique(sce$genotype))

pal <- pal[1:len]

names(pal) <- unique(sce$genotype)

col <- pal[match(sce$genotype, names(pal))]

plot(res@reducedDim, col = col, pch = 16, asp = 1, xlab = "Dimension 1", ylab = "Dimension 2")

lines(res, type = "lineages")

legend("topright", legend = unique(sce$genotype), col = pal, pch = 16)
```

### Cell types

Infer trajectory between all cell types using UMAP dimensions:

```{r}
sce <- slingshot(sce, clusterLabels = "cellassign_celltype", reducedDim = "UMAP")
```

Extract slingshot results from experiment data:

```{r}
res <- SlingshotDataSet(sce)
```

Inspect UMAP plot coloured by cell type:

```{r}
pal <- RColorBrewer::brewer.pal(9, name = "Set1")

len <- length(unique(sce$cellassign_celltype))

pal <- pal[1:len]

names(pal) <- unique(sce$cellassign_celltype)

col <- pal[match(sce$cellassign_celltype, names(pal))]

plot(res@reducedDim, col = col, pch = 16, asp = 1, xlab = "Dimension 1", ylab = "Dimension 2")

lines(res, type = "lineages")

legend("topright", legend = unique(sce$cellassign_celltype), col = pal, pch = 16)
```

Inspect UMAP plot coloured by genotype:

```{r}
pal <- RColorBrewer::brewer.pal(9, name = "Set1")

len <- length(unique(sce$genotype))

pal <- pal[1:len]

names(pal) <- unique(sce$genotype)

col <- pal[match(sce$genotype, names(pal))]

plot(res@reducedDim, col = col, pch = 16, asp = 1, xlab = "Dimension 1", ylab = "Dimension 2")

lines(res, type = "lineages")

legend("topright", legend = unique(sce$genotype), col = pal, pch = 16)
```

## Summary

### Output

```{r output}
readr::write_rds(sce, here::here("data/trajectory.Rds"))
```

### Session

```{r session}
devtools::session_info()
```
